Index measures for oak decline severity using phenotypic descriptors

### 2020-03-03 manuscript/manuscript.Rmd
added detail on armillaria bleeds to introduction

Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011; @petr2014; @brown2018].
Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both long-term predisposing and one-off inciting factors.
Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
Over time, as lesion development continues beneath the bark, external cracks develop between the bark plates that are surrounded by blackening stains.
Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of calloused tissue on the exterior [@denman2014].

Oak decline syndromes are wide ranging in cause and attempts have been made to subdivide the types of decline associated with *Quercus robur* and *Q. petraea* based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
Mortality in trees suffering from COD can occur over a period of decades whereas AOD can occur in 5-10 years [@denman2009].
A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD.

@@

Assigning discrete decline status labels neglects the true spectrum of decline severity encountered in the field.
There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline severity spectrum [@pontius2014].
This has the potential to greatly improve monitoring, enable the early detection of decline for accurate forecasting and inform forestry management practices.
Consistent long term monitoring has been highlighted as a key requirement for understanding the threats to oak and safeguarding its future [@hartmann2018].
Coupled with this, monitoring methods should take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.

>>>>

Oak decline is expected to become increasingly prevalent in the UK and across Europe with the onset of climate change [@sturrock2011; @petr2014; @brown2018].
Oak declines are complex syndromes that are facilitated by a variety of both abiotic and biotic factors.
Abiotic factors can include both long-term predisposing and one-off inciting factors.
Biotic agents can include insect infestation such as the larvae of bark boring beetle *Agrilus biguttatus* or bacterial stem bleeds associated with Acute Oak Decline (AOD) [@brown2017a; @denman2014], as well as a variety of fungal root pathogens associated Chronic Oak Decline (COD) [@denman2017].
The bacterial stem bleeds associated with AOD initially start as active dark weeping patches.
Over time, as lesion development continues beneath the bark, external cracks develop between the bark plates that are surrounded by blackening stains.
Evidence of stem tissue repair can be seen as lignified-callus-like tissue that, when extensive, can fully occlude the internal lesion, appearing as exposed rolls of calloused tissue on the exterior [@denman2014].
Stem bleeds can also, but not always, be associated with *Armillaria* attack on the inner stem tissues of oak.
Active stem weeping and black staining can be found along the leading edge of the mycelial front of affected stem tissues [@denman2017].

Oak decline syndromes are wide ranging in cause and attempts have been made to subdivide the types of decline associated with *Quercus robur* and *Q. petraea* based on the duration over which a tree has been declining and the biotic agents associated with a declining tree.
Mortality in trees suffering from COD can occur over a period of decades whereas AOD can occur in 5-10 years [@denman2009].
A chronically declining oak can also subsequently begin to develop the stem bleeds associated with AOD [@denman2017].

@@

Assigning discrete decline status classes neglects the true spectrum of decline severity encountered in the field.
There is a growing need for sensitive, continuous measures of tree decline that accurately reflect the decline severity spectrum [@pontius2014].
This has the potential to greatly improve monitoring, enable the early detection of decline for accurate forecasting and inform forestry management practices.
Consistent long term monitoring has been highlighted as a key requirement for understanding the threats to oak and safeguarding its future [@hartmann2018].
Coupled with this, monitoring methods should take advantage of modern data analysis techniques whilst still retaining practical simplicity in the field.
